A Study of Residual Adapters for Multi-Domain Neural Machine Translation

### 2020-08-14 2020/wmt_20/emnlp2020.tex
Reviewed new section 2

In this context, our main contribution is a thorough experimental study of the use of residual adapters in a multi-domain translation adaptation. We notably explore ways to adjust and/or regularize adapter modules to handle situations where the adaptation data is very small. We also propose and contrast two new variants of the residual architecture: in the first one (highway residual adapters), adaptation still affects each layer of the architecture, but its effect is delayed till the last layer, thus making the architecture more modular and adaptive; in the second one (gated residual adapters), we explore ways to improve the performance in the face of train-test data mismatch. We experiment with two language pairs and report results that demonstrate the flexibility and effectiveness of these architectures. 
\fyDone{One bit of a conclusion here}\fyFuture{Build a proper training scenario for these two conditions}

@@

\subsection{Architecture \label{ssec:architecture}}
\fyTodo{More contexts and notations from the transformer}\fyTodo{Encoder / decoder layers}
For $i^{\text{th}}$layer of the encoder or the decoder, let $h^i$ be the output of the layer, of dimension $d$. The $i^{\text{th}}$ residual adapter modifies $h^i$ by applying several transformations as follows:\fyDone{Use align env}
\begin{align*}
  h^{i}_1 &= \mathbf{W}_{db}^{i}h^{i} + b^i_{1} \\
  h^{i}_2 &= \mathbf{RELU}(h_1^{i}) \\
  h^{i}_3 &= \mathbf{W}_{bd}^{i}h_2^{i} + b^i_{2} \\
  \bar{h}^{i} &= h^{i}_3 + h^i.
\end{align*}
Matrices $\displaystyle{\mathbf{W}_{db}^{i}\in\mathbb{R}^{d\times b}}$,$\displaystyle{\mathbf{W}_{bd}^{i}\in\mathbb{R}^{b\times d}}$, bias vectors $\displaystyle{b^i_{1} \in \mathbb{R}^{b}}$, $\displaystyle{b^i_{2} \in \mathbb{R}^{d}}$, b is a bottleneck dimension. For the sake of brevity, we denote $h^i_3 = \operatorname{ADAP}^{(i)}(h^i)$, and $\theta_{\operatorname{ADAP}^{(i)}}$ the set of parameters for the $i^{\text{th}}$ adapters.\fyDone{or is it $h_i$ ?}\fyDone{attention aux matrices $W_i$}

The "adapted" hidden vectors $\bar{h}^i_{ 1\leq i \leq n-1}$, where n is the number of layers, is the input of the $(i+1)^{\text{th}}$ \fyDone{Self attention ?} layer, $\bar{h}^n$ is passed to the decoder if it belongs to the encoder side, or the input of output layer if it belongs to the decoder side. In our architecture, there are 12 residual adapters, one for each of the six attention layers of the encoder and similarly for the decoder.\footnote{In the decoder, the stack of self-attention layers and cross encoder-decoder attention only counts as one attention layer and serves only one residual adapter} 
\fyTodo{Many question opens: how to train ? how many adapters ? which size ? etc}

The size of available corpora for each domain is extremely varying. For example, the corpus of domain \domain{tourism} in En-De contains only around $7\times10^3$ sentence pairs while the corpus of domain \domain{news} in En-De contains approximately $3\times10^6$ sentence pairs. For extremely small domains, we could economize the number of parameters by reducing the number of residual adapters in the architecture. While only a limited number of adapters are incorporated to the NMT model, which positions in the model are more important in the domain adaptation task? We would like to analyze the impact of position and number of residual adapters involved in the adapted model. It is conjectured that the higher layers represent more global patterns such as semantic while the lower layers represent more local patterns such as syntactic \cite{raganato18analysis}\fyTodo{Missing reference here}. The domain shift in local patterns and global patterns has not yet studied. In this paper, we do not intend to study this aspect of the domain adaptation problem. We would like to give an extensive comparison between domain adaptation in different levels in NMT model.

Fine-tuning on small corpora can still lead to an overfitted model even when the fine-tuning is applied to a few residual adapters. To avoid this situation, we have several options including: reducing the size of the adapter module, applying weight decay or layer regularization to the adapter. The first option is not flexible \fyTodo{compatible ???} if we consider the probable growth of in-domain data. The second and third choices are preferable to regularize a large adapter which will scale if the in-domain corpora extend .\fyTodo{Style}

Weight decay applies a penalization on the weights of the adapters.\fyDone{The second summation also runs over $x,y$ ? I think not}
\begin{equation}
  \begin{split}
    \bar{L} & = \frac{1}{\#(x,y)}\mathop{\sum}_{x,y}( - \log(P(y|x))) \\
    & + \lambda * \sum_{i \in \{1,..,6\} \otimes \{enc, dec\}} \normL{\theta_{\operatorname{ADAP}^{(i)}}}
  \end{split}
\end{equation}
Where $\theta_{\operatorname{ADAP}^{(i)}}$ are parameters of $i^{\text{th}}$ adapter

Layer regularization applies a penalization on the output of the adapters.
\begin{equation}
  \begin{split}
    \bar{L} & = \frac{1}{\#(x,y)}\mathop{\sum}_{x,y} (-\log(P(y|x)) \\
    & + \lambda * \sum_{i \in \{1,..,6\} \otimes \{enc, dec\}} \normL{\operatorname{ADAP}^{(i)}(h_i(x,y))})
  \end{split}
\end{equation}

In a multi-domain context, there are 2 ways to train domain-specific NMT model using adapters including pretraining generic NMT then finetuning adapters; or training both NMT model and adapters at once from the beginning using multi-task learning. In the literature of Machine Learning, \cite{Caruana97multitask} proposed an architecture suitable for multi-task learning. The corresponding network comprises a ground layer, which is shared between tasks and is followed by task-related layers. The network is then trained from scratch by fueling task-related examples in a batch and updating all parameters (shared and task-specific) at the same time. The multi-domain context is similar to the multi-task context if we consider a domain as a task\cite{Dredze08online}. Therefore, we can also use multi-task training to optimize the parameters of NMT model and the parameters of the residual adapters at once from the beginning. To assess the efficacy of multi-task training in the multi-domain context, we would like to compare the finetuning approach with the multi-task training approach. \fyTodo{Change the training regime}

\subsection{Highway Residual Adapters \label{ssec:highway}}
Finally, we would like to assess the performance of a variant of residual adapters in which the adapters go directly to the end of the stack of attention layers. We graphically demonstrate this version in Figure~\ref{fig:hrl-architecture}. In this version, only the last hidden vector of the encoder (decoder) is modified as follow
\begin{equation}
\bar{h}^n = h^n + \displaystyle{\mathop{\sum}_{1 \leq i \leq n} ADAP^i(h^i)}
\end{equation}
The benefit of this variant is that it allows us to reuse hidden vectors $h^i$ of all hidden layers while adapting to multiple domains. In the forward step, the hidden vectors $h^i$ are calculated once, then adapters take these vectors to modify the last layer. Therefore, we can finetune the model to multiple domains at once without recomputing $h^i$. The second minor benefit of this variant is that it allows implementing the Gated Residual Adaptaters version that will be described in the following section \ref{ssec:gate}.
\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.3]{fig/highway_residual}
  \caption{Highway multi-domain network with residual layers}
\label{fig:hrl-architecture}
\end{figure}
\subsection{Gated Residual Adaptaters \label{ssec:gate}}
\mpTodo{Formalizing problem, network design, training algorithm}
To translate a sentence, we have 2 options: using a domain-adapted model or using the generic model. In the case of choosing the wrong domain, the performance is dramatically worse than using the generic model. Therefore, we would like to design a version of applying residual adapter that allows us to automatically accept or reject given domain information, i.e, even if we choose a wrong adapter, the NMT model is still able to reject the change output from the adapter. In this version, we reuse the highway version of residual adapters. 
\fyTodo{Consistency of notations wrt section 2.1}
Given domain $k$ with set of adapters $ADAP^i_k$, in the model adapted to domain k, instead of using the formulation $\bar{h}^n = h^n + \displaystyle{\mathop{\sum}_{1 \leq i \leq n} ADAP_k^i(h^i)}$ (encoder or decoder's side), we propose "gated" version uses $\bar{h}^n = h^n + \displaystyle{\mathop{\sum}_{1 \leq i \leq n} ADAP_k^i(h^i) \odot{} z_k(h^n)}$. Where $z_k(h^n(w))$ is designed to reflect the relatedness of word $w$ to domain $k$: the more likely $w$ is in domain $k$, the larger $z_k(h^n(w))$ should be. The gated version aims to reduce the change $\displaystyle{\mathop{\sum}_{1 \leq i \leq n} ADAP_k^i(h^i)}$ added to $h^n$ for words that are not typical of domain $k$. In our implementation, we incorporate 2 domain classifiers on top of the encoder and the decoder that take the last hidden layer of encoder (decoder) as input, we take the posterior probability $P(d=k|w)$ as value of $z_k(h^n(w))$.

The training of Gated Residual Adaptaters comprises three main steps:
\begin{itemize}
	\item Pretraining a generic model with mixed corpora comprising data from multiple domains.
	\item Training a domain classifier on top of the encoder and decoder; during this step, the parameters of the generic model are frozen. This model computes the posterior domain probability $p(k|h(w_t))$ for each word $w_t$ based on the representation computed by the last layer.
	\item Training the parameters of adapters with in-domain data separately for each domain while freezing the parameters of the generic model and of the domain classifiers.
\end{itemize}

>>>>

In this context, our main contribution is a thorough experimental study of the use of residual adapters in multi-domain translation. We notably explore ways to adjust and/or regularize adapter modules to handle situations where the adaptation data is very small. We also propose and contrast two new variants of the residual architecture: in the first one (highway residual adapters), adaptation still affects each layer of the architecture, but its effect is delayed till the last layer, thus making the architecture more modular and adaptive; in the second one (gated residual adapters), we explore ways to improve the performance in the face of train-test data mismatch. We experiment with two language pairs and report results that demonstrate the flexibility and effectiveness of these architectures. 
\fyDone{One bit of a conclusion here}\fyFuture{Build a proper training scenario for these two conditions}

@@

\subsection{Basic architecture \label{ssec:architecture}}
\fyDone{More contexts and notations from the transformer}\fyDone{Encoder / decoder layers}

\subsubsection{The computation of adapter layers}
Our reference architecture is the Transformer model of \cite{Vaswani17attention}, which we assume contains a stack of layers both in the encoder and the decoder side. Each layer contains two subparts, an attention layer and a dense layer. Details may vary from one implementation to another, we simply content here that each layer $i \in \{1 \dots L\}$ (in the encoder or the decoder) computes a transform of a fixed-length sequence of $d$-dimensional input vectors $h^{i-1}$ into a sequence of output vectors $h^{i}$.

The residual adapter architecture extends this model by adding one new computational module for each layer. The $i^{\text{th}}$ such residual adapter thus modifies $h^i$ by computing the following transformations:\fyDone{Use align env}
\begin{align*}
  h^{i}_1 &= \mathbf{W}_{db}^{i}h^{i} + b^i_{1} \\
  h^{i}_2 &= \mathbf{ReLU}(h_1^{i}) \\
  h^{i}_3 &= \mathbf{W}_{bd}^{i}h_2^{i} + b^i_{2} \\
  \bar{h}^{i} &= h^{i}_3 + h^i.
\end{align*}
Overall, the  $i^{\text{th}}$ adapter is thus parameterized by matrices $\displaystyle{\mathbf{W}_{db}^{i}\in\mathbb{R}^{d\times b}}$,$\displaystyle{\mathbf{W}_{bd}^{i}\in\mathbb{R}^{b\times d}}$, bias vectors $\displaystyle{b^i_{1} \in \mathbb{R}^{b}}$, $\displaystyle{b^i_{2} \in \mathbb{R}^{d}}$, with $b$ the dimension of the adapter (with $d \gg b$)\fyTodo{Check this}. For the sake of brevity, we will simply denote $h^{i}_3 = \operatorname{ADAP}^{(i)}(h^i)$, and $\theta_{\operatorname{ADAP}^{(i)}}$ the corresponding set of parameters.\fyDone{or is it $h_i$ ?}\fyDone{attention aux matrices $W_i$}

The "adapted" hidden vectors $\bar{h}^i_{ 1\leq i \leq L-1}$, where $L$ is the number of layers, will then be the input of the $(i+1)^{\text{th}}$\fyDone{Self attention ?} layer; $\bar{h}^L$ is passed to the decoder if it belongs to the encoder side, or is the input of output layer if it belongs to the decoder side. Note that zeroing out all adapters enables us to recover the basic Transformer, with $\bar{h}^{i} = h^i$ for all $i$.

In the experiments reported in Section~\ref{sec:exp}, we will use $2\times{}L=12$ residual adapters, one for each of the $L=6$ attention layers of the encoder and similarly for the decoder.\footnote{In the decoder, the stack of self-attention layers and cross encoder-decoder attention only counts as one attention layer and only serves one residual adapter.}

\subsubsection{Design space and variants}
This general architecture leaves open many design choices pertaining to the details of the network organization, the training procedure and the corresponding objective function.

A first question concerns the number of adapter layers. While in principle, all Transformer layers can be subject to adaptation, it is nonetheless worthwhile to consider simpler adaptation schemes, which would only alter a limited number of layers. Such strategy might be especially relevant when the training corpus contains very small domains, such as the ones considered in our experiments, and for which a complete adaptation may not be necessary or/and or prone to overfitting. Likewise, it might be meaningful to explore ways to share subsets of adapters across domains. This in turn raises the issues of which layer(s) to adapt, a question that can be approached in the light of recent analyses of the Transformer architecture which conjectured that the higher layers encode more global patterns with a more ``semantic'' interpretation, while the lower layers encode more local patterns akin to morpho-syntactic information \cite{raganato18analysis}.

% The size of available corpora for each domain is extremely varying. For example, the corpus of domain \domain{tourism} in En-De contains only around $7\times10^3$ sentence pairs while the corpus of domain \domain{news} in En-De contains approximately $3\times10^6$ sentence pairs. For extremely small domains, we could economize the number of parameters by reducing the number of residual adapters in the architecture. While only a limited number of adapters are incorporated to the NMT model, which positions in the model are more important in the domain adaptation task? We would like to analyze the impact of position and number of residual adapters involved in the adapted model. It is conjectured that the higher layers represent more global patterns such as semantic while the lower layers represent more local patterns such as syntactic \cite{raganato18analysis}\fyTodo{Missing reference here}. The domain shift in local patterns and global patterns has not yet studied. In this paper, we do not intend to study this aspect of the domain adaptation problem. We would like to give an extensive comparison between domain adaptation in different levels in NMT model.

A related question concerns the regularization of adapter layers to mitigate overfitting. Reducing the number of adapters, or their dimensions, is simple, but these choices are difficult to optimize numerically -- an issue that becomes important as the number of domain grows. Less naïve alternatives can also be entertained, such as applying weight decay or layer regularization to the adapter. Implementing these requires to modify the objective function in a way that still allows for a smooth optimization problem. For instance, weight decay applies a penalization on the weights of the adapters, complementing the cross-entropy term with a function of the norm of the parameters: \fyDone{The second summation also runs over $x,y$ ? I think not}
\begin{equation*}
  \begin{split}
    \bar{L} & = \frac{1}{\#(x,y)}\mathop{\sum}_{x,y}( - \log(P(y|x))) \\
    & + \lambda  \sum_{i \in \{1,..,6\} \otimes \{enc, dec\}} \normL{\theta_{\operatorname{ADAP}^{(i)}}}
  \end{split}
\end{equation*}
An alternative scheme is \emph{layer regularization}, which penalizes the output of the adapters, corresponding to the following new training objective:
\begin{equation*}
  \begin{split}
    \bar{L} & = \frac{1}{\#(x,y)}\mathop{\sum}_{x,y} (-\log(P(y|x)) \\
    & + \lambda \sum_{i \in \{1,..,6\} \otimes \{enc, dec\}} \normL{\operatorname{ADAP}^{(i)}(h_i(x,y))})
  \end{split}
\end{equation*}

Finally, another independant design choice relates to the training strategy for adapters. A first option is to generalize supervised domain adaptation to multi-domain adaptation and to proceed in two steps: (a) train a generic model with all the available data; (b) train eacj adapter layer with domain-specific data, keeping the generic model parameters unchanged. Another strategy is to adopt the view of \citet{Dredze08online}, where the multi-domain setting is viewed as an instance of multi-task learning \cite{Caruana97multitask}, where each domain corresponds to a specific task. This suggest to train the models as it were in a multi-task mode, training all the parameters from scratch: the generic parameters values will still depend on all the available data, while each adapter will only be trained with the corresponding in-domain data.

\subsection{Highway Residual Adapters \label{ssec:highway}}

In the basic architecture described in Section~\ref{ssec:architecture}, the computation performed by lower level layers will impact all the subsequent layers. In this section, we introduce an alternative implementation of the same idea, which however delays the adaptation of each layer to the last layer (of the encoder or the decoder) as represented on Figure~\ref{fig:hrl-architecture}. While the basic architecture performs adaptation in sequence, we propose here to perform it in parallel. This means that in this version, only the last hidden vector of the encoder (decoder) is modified according to:
\begin{equation}
  \bar{h}^L = h^L + \displaystyle{\mathop{\sum}_{1 \leq i \leq L} ADAP^i(h^i)} \label{eq:highway-output}
\end{equation}

One obvious benefit of this variant is that it allows us to reuse the hidden vectors $h^i$ of all hidden layers when computing an adapted output for several domains during the inference. In this situation, the forward step needs only to compute the hidden vectors $h^i$ once for the inner encoder layers, before an adapted sequence of vectors is computed at the topmost layer. Therefore, we can finetune the model to multiple domains at once without recomputing $h^i$. This variant also opens the way to more parameter sharing across adapters, a perspective that we will not explore further in this work. Instead, we use it to develop a second variation of the adapter model, that is presented in the next section.

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.3]{fig/highway_residual}
  \caption{Highway multi-domain network with residual layers}
  \label{fig:hrl-architecture}
\end{figure}

\subsection{Gated Residual Adapters \label{ssec:gate}}
\mpTodo{Formalizing problem, network design, training algorithm}
The basic architecture presented above rests on a rather simplistic view of ``domains'' as made of well-separated and unrelated pieces of texts that are processed independently during the adaptation step. Likewise, when translating test documents, we need to make a choice between either using a specific domain-adapted model or resorting to the generic, unadapted model. In this context, errors on the domain label can have a strong (negative) impact on the trained model, or on the translation performance.

Therefore, we would like to design a version of applying residual adapter that is more robust to such domains error. This variant, called the \emph{gated residual adapter models}, relies on the training of a supplementary trained component that will help decide whether to activate, on a word per word basis, a given residual layer, and to regulate the strength of this activation. To this end , we extend the highway version of residual adapters as follows.
\fyDone{Consistency of notations wrt section 2.1}

Formally, we replace the adapter computation of equation~\eqref{eq:highway-output} and take the adapted hidden (topmost) layer to be computed as (this is for domain $k$):
\begin{equation}
  \bar{h}^L = h^L + \displaystyle{\mathop{\sum}_{1 \leq i \leq L} \operatorname{ADAP}_k^i(h^i) \odot{} z_k(h^L)}, \label{eq:gated-output}
\end{equation}
where $z_k(h^L(w))$ measures the relatedness of word $w$ to the domain $k$. The more likely $w$ is in domain $k$, the larger $z_k(h^n(w))$ should be; conversely, for words that are not typical of any domain $k$ (eg.\ function words),  adaptation is minimum and the adapted corresponding encoder output will remain close to the output of the generic model ($h^L$). In our implementation, we incorporate two domain classifiers on top of the encoder and the decoder that take the last hidden layer of encoder (resp.\ decoder) as input, and use the posterior probability $P(d=k|w)$ as the value of $z_k(h^n(w))$.

The training of gated residual adapters thus comprises three main steps:
\begin{itemize}
\item Learn a generic model with mixed corpora comprising data from multiple domains.
\item Train a domain classifier on top of the encoder and decoder; during this step, the parameters of the generic model are frozen. This model computes the posterior domain probability $p(k|h(w_t))$ for each word $w_t$ based on the representation computed by the last layer.
\item Train the parameters of adapters with in-domain data separately for each domain, while freezing the generic model and the domain classifiers.
\end{itemize}
